---
title: "Historical evolution of NOAA records"
author: "David Conejero"
date: "26 de julio de 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warnings = FALSE)
knitr::opts_chunk$set(fig.path = "figure/")
```

## Synopsis

As part of the analysis on the [Most harmful weather events in USA][2] I detected
two possible sources of errors on the analysis of the data provided by the National Oceanic 
and Atmospheric Administration (NOAA) [Storm Database][1].

One was related to the fact that there were some EVTYPE errors.
The other was that, since the information is recorded from 1950's, the precision
on the data registration has increased along the years, and some bias could be introduced
to the analysis done using the database.

This report includes some information that may help work properly with the data provided by the database.

## Data Processing

First of all we upload the libraries we are going to use

```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
```

We get the information from the StormDatabase and we upload it to the system. It 
contains weather events from 1950 up until today.

```{r load, cache=TRUE}
destfile="stormdata.csv.bz2"
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                      destfile)
stormdata<-read.csv(destfile)
```


## Weather events evolution

On this report I want to check how the registration of events evolved along the years.

One of the first things to be noted is that there are too many types of events 
registered (985). And some of them are erroneous or duplicated

```{r events}
weather_types<-levels(stormdata$EV)
length(weather_types)
set.seed(400)
weather_types[sample(1:980,30)]
```

However, most of them appear just once or twice, and they don't have a big impact
on the data, and no impact at all on the analysis we are doing right now.

In order to analyze the historical evolution of "well" registered events, I will
group the data by events, and choose the top 5 most repeated event types along history.
For that, previously I would have assigned one year to each event. In this case
I chose the year when the event begins (since it is not expected that one event lasts
for two different years, except on the specific event that it started on Dec 31st and
ended January 1st. In any case, it doesn't matter on which year this event is registered
if it is only registered once)

```{r historical, fig.width=10}
# This step will allow to simplify the data.frame used
historical<-select(stormdata, BGN_DATE, EVTYPE, FATALITIES, INJURIES)
historical<-mutate(historical, year=year(strptime(as.character(BGN_DATE),"%m/%d/%Y %H:%M:%S")))
by_event<-stormdata %>% 
          group_by(EVTYPE) %>% 
          summarise(c=n())
top_types<-by_event[order(by_event$c,decreasing = T),]
tt<-head(top_types$EVTYPE, 5)
tt_historical<-filter(historical, EVTYPE %in% tt)
#checking
tt_historical %>% group_by(EVTYPE) %>% summarise(c=n()) %>% arrange(desc(c))


count_tt_historical<- tt_historical %>% group_by(EVTYPE, year) %>% summarise(c=n())

ggplot(data=count_tt_historical, aes(x=year,y=c,col=EVTYPE)) +geom_point()
```

As it can be seen on the figures, while the number of TORNADOs reported is more 
or less the same from 50's to nowadays, HAIL and TSTM WIND is growing exponentially
from 80's, there are no reports of FLASH FOOD before 90's, and there is a lot of
variance in the distribution of THUNDERSTORM WIND reports.

To check the evolution of the elements more harmful for population and economics,
we should get the list of events and see its evolution. To see where top_fatalities,
top_injuries, top_prop, top_crop and top_total are obtained, please, check [Most harmful weather events in USA][2]

```{r historical_harmful, fig.width=10}

by_event<-stormdata %>% 
          group_by(EVTYPE) %>% 
          summarise(sum_fatalities=sum(FATALITIES), 
                    sum_injuries=sum(INJURIES))


top_fatalities<-by_event %>% arrange(desc(sum_fatalities)) %>% head(5)
top_injuries<-by_event %>% arrange(desc(sum_injuries)) %>% head(5)

economic<-select(stormdata, BGN_DATE, EVTYPE, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP)

economic<-mutate(stormdata, 
                 PROP=case_when(PROPDMGEXP=="M" ~ PROPDMG*10^6, 
                                PROPDMGEXP=="B" ~ PROPDMG*10^9, 
                                PROPDMGEXP=="K" ~ PROPDMG*10^3, 
                                TRUE ~ 0),
                 CROP=case_when(CROPDMGEXP=="M" ~ CROPDMG*10^6, 
                                CROPDMGEXP=="B" ~ CROPDMG*10^9, 
                                CROPDMGEXP=="K" ~ CROPDMG*10^3, 
                                TRUE ~ 0),
                 TOTAL=PROP+CROP
                 )

e_by_event<-economic %>% 
            group_by(EVTYPE) %>% 
            summarise(sum_prop=sum(PROP), 
                      sum_crop=sum(CROP),
                      sum_total=sum(TOTAL))

top_prop<-e_by_event %>% arrange(desc(sum_prop)) %>% head(5)
top_crop<-e_by_event %>% arrange(desc(sum_crop)) %>% head(5)
top_total<-e_by_event %>% arrange(desc(sum_total)) %>% head(10)

tt<-unique(c(as.character(top_fatalities$EVTYPE),as.character(top_injuries$EVTYPE),as.character(top_prop$EVTYPE),as.character(top_crop$EVTYPE),as.character(top_total$EVTYPE)))

tt_historical<-filter(historical, EVTYPE %in% tt)

count_tt_historical<- tt_historical %>% group_by(EVTYPE, year) %>% summarise(c=n())

ggplot(data=count_tt_historical, aes(x=year,y=c,col=EVTYPE)) +geom_point()
```

As it can be seen, almost all the events are reported after 1990, except for TORNADO,
TSTM WIND and HAIL. The explanation of the way they have been registered has already been done on the previous figure.

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf 
[2]: https://rpubs.com/aldacon/reproducible-research-2
