---
title: "Most harmful weather events in USA"
author: "David Conejero"
date: "26 de julio de 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warnings = FALSE)
knitr::opts_chunk$set(fig.path = "figure/")
```

## Synopsis

In this report we aim to provide information on which is the more harmful weather
event in USA. In order to do that we used the National Oceanic 
and Atmospheric Administration (NOAA) [Storm Database][1]. This database is an official
NOAA publication which documents the occurrence of storms and other significant weather phenomena having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce. 

In this report we check the impact of weather events from two different visions: 
harm done to the population health and economic impact. From those analyses, we could
see that the most harmful events for the population are Tornadoes and Flood have the biggest
economic impact.


## Data Processing

First of all we upload the libraries we are going to use

```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
```

We get the information from the StormDatabase and we upload it to the system. It 
contains weather events from 1950 up until today.

```{r load, cache=TRUE}
destfile="stormdata.csv.bz2"
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                      destfile)
stormdata<-read.csv(destfile)
```

We checked that the information is properly formatted

```{r check}
str(stormdata)
```

EVTYPE contains the weather event name. But there are 980 *different* weather events.
We took a closer look:

```{r events}
weather_types<-levels(stormdata$EV)
set.seed(400)
weather_types[sample(1:980,30)]
```

As it can be seen there are strange texts like "Summary August 7" and *duplicated*
entries like "HAIL 1.00", "HAIL 1.75)", and we know there is one "HAIL". In order
to treat this properly a more in depth study should be done on these values and 
some data should be cleaned up. We can check how many of these events may have real
impact, either in the number of events or fatalities registered:

```{r impact}
by_event<-stormdata %>% 
          group_by(EVTYPE) %>% 
          summarise(sum_fatalities=sum(FATALITIES), 
                    c=n())

quantile(by_event$c, c(0.8,0.9,0.95,0.99,1))
sum(head(by_event$c[order(by_event$c,decreasing = T)],30))/sum(by_event$c)

quantile(by_event$sum_fatalities, c(0.8,0.9,0.95,0.99,1))
sum(head(by_event$sum_fatalities[order(by_event$sum_fatalities,decreasing = T)],30))/sum(by_event$sum_fatalities)
```
As it can be seen, the impact in both, fatalities and number of events, is very low
for most of them. And since this is just scholar assignment, I will let the analysis for another time. However, we will see it will have some impact on the classification 
of events based on the overall economical impact.

Another aspect to be considered is that, from the documentation we know that the way the information is registered has been changing along the years. So I decided to check how may have impacted this changes on the registration process and how may it impact on the figures of this report. On this [report][2] there is a description of this analysis. But it doesn't seem to change significantly the results found on this report.

Taking into consideration these two factors, we start the analysis on the impact on population health and economic consequences.

## Results

### Harm done to population

There may be 2 different kinds of harms done to the population: fatalities and injuries.

We will analyze both separately. First we consider fatalities caused by weather events:

#### Fatalities

```{r fatalities}
by_event<-stormdata %>% 
          group_by(EVTYPE) %>% 
          summarise(sum_fatalities=sum(FATALITIES), 
                    sum_injuries=sum(INJURIES))

top_fatalities<-by_event %>% arrange(desc(sum_fatalities)) %>% head(5)
top_fatalities$EVTYPE<-factor(top_fatalities$EVTYPE, 
                              levels=top_fatalities$EVTYPE[order(top_fatalities$sum_fatalities,
                              decreasing = TRUE)])
```

From this we can see that the events with more fatalities are:

```{r fatalities_plot, echo=FALSE}
ggplot(top_fatalities, aes(x=EVTYPE, y=sum_fatalities)) + 
        geom_bar(stat="identity", fill=c("red")) + labs(x="Events", y="Fatalities")
```

And the most harmful event is the TORNADO with
```{r top_fatalities}
top_fatalities$sum_fatalities[1]
```

#### Injuries

```{r injuries}
top_injuries<-by_event %>% arrange(desc(sum_injuries)) %>% head(5)
top_injuries$EVTYPE<-factor(top_injuries$EVTYPE,
                            levels=top_injuries$EVTYPE[order(top_injuries$sum_injuries,
                            decreasing = TRUE)])
```

From this we can see that the events with more injuries are:

```{r injuries_plot, echo=FALSE}
ggplot(top_injuries, aes(x=EVTYPE, y=sum_injuries)) + 
        geom_bar(stat="identity", fill=c("red")) + labs(x="Events", y="Injuries")
```

And the event with more injured people is, again, the TORNADO with
```{r top_injuries}
top_injuries$sum_injuries[1]
```

### Economic impact

On the database, the economic impact is computed by two different fields: property
damage and crop damage. We will follow the same approximation, but at the end we will
sum up all the economical damage to see the impact of the different weather events
in US economy.

On the database, the information for property and crop damage is splitted in 2
columns, one containing the *value* of the economic loss, the other the *magnitude
order*. However, the *magnitude order* has some errors:


```{r magnitud_order_errors}
table(stormdata$PROPDMGEXP)
table(stormdata$CROPDMGEXP)
```

As it can be seen, most of the values are K (thousand) or M (millions),
from the documentation we know that B are billions (10^9), but we don't know
what the other values mean. Since they are very few (compared to the others) I 
will not take them into consideration at all. However, it must be noted that, 
depending on its **real** value (8->10^8?), they can have a *huge* impact on this analysis.

First we extract the economic cost for property damage, crop damage and total by
type of event.

```{r economic}
economic<-select(stormdata, BGN_DATE, EVTYPE, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP)

economic<-mutate(stormdata, 
                 PROP=case_when(PROPDMGEXP=="M" ~ PROPDMG*10^6, 
                                PROPDMGEXP=="B" ~ PROPDMG*10^9, 
                                PROPDMGEXP=="K" ~ PROPDMG*10^3, 
                                TRUE ~ 0),
                 CROP=case_when(CROPDMGEXP=="M" ~ CROPDMG*10^6, 
                                CROPDMGEXP=="B" ~ CROPDMG*10^9, 
                                CROPDMGEXP=="K" ~ CROPDMG*10^3, 
                                TRUE ~ 0),
                 TOTAL=PROP+CROP
                 )

e_by_event<-economic %>% 
            group_by(EVTYPE) %>% 
            summarise(sum_prop=sum(PROP), 
                      sum_crop=sum(CROP),
                      sum_total=sum(TOTAL))
```

#### Property damage

First we get the property damage cost by type of weather event.

```{r cost_in_property_damage}
top_prop<-e_by_event %>% arrange(desc(sum_prop)) %>% head(5)
top_prop$EVTYPE<-factor(top_prop$EVTYPE, 
                        levels=top_prop$EVTYPE[order(top_prop$sum_prop,
                                                     decreasing = TRUE)])

```

From this we can see that the events with greatest property damages are:

```{r top_property_damages, echo=FALSE}
select(top_prop,EVTYPE,sum_prop)
```

And the most costly event for properties is the FLOOD with almost 150B$

#### Crop damage

First we get the crop damage cost by type of weather event.

```{r cost_in_crop_damage}
top_crop<-e_by_event %>% arrange(desc(sum_crop)) %>% head(5)
top_crop$EVTYPE<-factor(top_crop$EVTYPE, 
                        levels=top_crop$EVTYPE[order(top_crop$sum_crop,
                                                     decreasing = TRUE)])
```

From this we can see that the events with greatest crop damages are:

```{r top_crop_damages, echo=FALSE}
select(top_crop,EVTYPE,sum_crop)
```

And the most costly event for crop is the DOUGHT with almost 14B$

#### Total damage

First we get the total damage cost by type of weather event.

```{r cost_in_total}
top_total<-e_by_event %>% arrange(desc(sum_total)) %>% head(10)
top_total$EVTYPE<-factor(top_total$EVTYPE, 
                        levels=top_total$EVTYPE[order(top_total$sum_total,
                                                     decreasing = TRUE)])
```

From this we can see that the events with greatest economic damages are:

```{r top_total, echo=FALSE}
select(top_total,EVTYPE,sum_total)
```

We can also see it if we plot those values in the following figures:
```{r plot, fig.width=10}

plot_p<-ggplot(top_prop, aes(x=EVTYPE, y=sum_prop/10^9)) + 
        geom_bar(stat="identity", fill=c("red"))  + 
        labs(y="Property damages (Billion $)", x="Event") + 
        theme(axis.text.x = element_text(size=7))

plot_c<-ggplot(top_crop, aes(x=EVTYPE, y=sum_crop/10^9)) + 
        geom_bar(stat="identity", fill=c("red"))  + 
        labs(y="Crop damages (Billion $)", x="Event")


plot_t<-ggplot(top_total, aes(x=EVTYPE, y=sum_total/10^9)) + 
        geom_bar(stat="identity", fill=c("red"))  + 
        labs(y="Economic damage (Billion $)", x="Event") + 
        theme(axis.text.x = element_text(size=8))

grid.arrange(plot_t,plot_p,plot_c, ncol=2, layout_matrix=cbind(c(1,2),c(1,3)))
```


And the most costly event is the FLOOD with more than 150B$, while DROUGHT is
only the 7th, which it seems to make sense, since it has almost no economical 
impact on property damage. 

One think that needs to be pointed out is that in this list, we can see 
"HURRICANE/TYPHOON" and "HURRICANE". It needs a further analysis to understand 
if this distinction is needed or it can all the events been labeled just as 
"HURRICANE".

## Conclusions

The most harmful weather events for population are Tornadoes, with 5,633 fatalities
and more than 9,000 injuries.

The most harmful weather events for economy are Floods, with an impact of more 
than 150B$ in property damages. However, for crops, the most harmful economic 
impact is from the DROUGHT with 15B$.


[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf 
[2]: https://rpubs.com/aldacon/historical-analysis